---
title: "variant_annotation.rmd"
output: html_document
date: "2023-05-30"
---

```{r}
library(stringr)
library(dplyr)
library(RColorBrewer)
library(maftools)
library(ggplot2)
library("BSgenome.Hsapiens.UCSC.hg38", quietly = TRUE)
library('NMF')
library(sigminer)
library(VariantAnnotation)

res_path <- "../results/combined_batches_wes/"
```

# Loading in MAF files

```{r}
maf_files = c()

for (maf_file in list.files("/media/gedac/kane/projects/booney_wes_clean/results/first_batch_wes/maf",
                            full.names=TRUE)){
  mymaf = read.maf(maf=maf_file)
  maf_files = append(maf_files, mymaf)
}

for (maf_file in list.files("/media/gedac/kane/projects/booney_wes_clean/results/second_batch_wes/maf",
                            full.names=TRUE)){
  mymaf = read.maf(maf=maf_file)
  maf_files = append(maf_files, mymaf)
}

merged_maf <- maftools::merge_mafs(maf_files)

# Saving maf file
write.mafSummary(merged_maf, basename=paste0(res_path, "merged_mafs_allBatches_160124"))
# Creating the patient and sample column for better annotation
#write.mafSummary(merged_maf, basename="merged_mafs_190923")
```
# Loading MAF file (full) and metadata

```{r}
merged_maf <- read.maf(paste0(res_path,"merged_mafs_allBatches_160124_maftools.maf"), 
                 verbose=TRUE)

# Load in metadata containing KRAS mutation information from sanger seuqencing
metadata_wes <- readxl::read_excel("../metadata/metadata_all_samples.xlsx")

# Cleaning the clinical data column to match with metadata spec
merged_maf@clinical.data <- merged_maf@clinical.data %>%
  tidyr::separate(Tumor_Sample_Barcode, into = c("Patient", "Sample", "Treatment"), sep = "_", remove=FALSE) %>%
  dplyr::mutate(Sample_Label = stringr::str_extract(Sample, "[A-Za-z]+")) %>%
  dplyr::mutate(Sample_Clean = stringr::str_replace(Sample, "RT.*", "R")) %>%
  data.table::setDT()

# Adding in metadata info
merged_maf@clinical.data <- dplyr::left_join(merged_maf@clinical.data, metadata_wes, by=c("Sample_Clean" = "Label")) %>%
  dplyr::rename(Stage = `Stage at diagnosis...10`)

# Subsetting for tissue vs organoid samples
tissue_samples <- merged_maf@clinical.data$Tumor_Sample_Barcode[grep(merged_maf@clinical.data$Tumor_Sample_Barcode, pattern="*TISSUE$")]
organoid_samples <- merged_maf@clinical.data$Tumor_Sample_Barcode[-grep(merged_maf@clinical.data$Tumor_Sample_Barcode, pattern="*TISSUE$")]

merged_maf_tissue_only <- subsetMaf(merged_maf, tsb= tissue_samples)
merged_maf_organoid_only <- subsetMaf(merged_maf, tsb= organoid_samples)
```
# Export the relevant stats for mutational information for investigators

```{r}
merged_maf@data %>% dplyr::mutate(patient_id = str_extract(Tumor_Sample_Barcode, "^PCA[0-9]+")) %>%
  dplyr::select(Tumor_Sample_Barcode, patient_id, Hugo_Symbol, HGVSp_Short, Exon_Number, t_ref_count, t_alt_count, t_depth,Consequence, SIFT, PolyPhen, IMPACT) %>% dplyr::group_by(Hugo_Symbol) %>% write.csv(paste0(res_path,"full_mutation_list.csv"))
```

#. Checking KRAS mutation status of samples 

```{r}
merged_maf@data %>% dplyr::filter(Hugo_Symbol == "KRAS") %>% 
  dplyr::select(Tumor_Sample_Barcode, HGVSp_Short, t_ref_count, t_alt_count) %>%
  tidyr::separate(Tumor_Sample_Barcode, into = c("Patient", "Sample", "Treatment"), sep = "_", remove=FALSE) %>%
  dplyr::mutate(Sample_Label = stringr::str_extract(Sample, "[A-Za-z]+"))  %>%
  dplyr::mutate(Sample_Clean = stringr::str_replace(Sample, "RT.*", "R")) 

sanger_kras <- merged_maf@clinical.data %>%
  dplyr::select(Sample_Clean, KRAS) %>%
  dplyr::distinct() %>%
  dplyr::mutate(Source = "Sanger")

wes_kras <- merged_maf@data  %>% dplyr::filter(Hugo_Symbol == "KRAS") %>% 
  dplyr::select(Tumor_Sample_Barcode, HGVSp_Short, t_ref_count, t_alt_count) %>%
  tidyr::separate(Tumor_Sample_Barcode, into = c("Patient", "Sample", "Treatment"), sep = "_", remove=FALSE) %>%
  dplyr::mutate(Sample_Label = stringr::str_extract(Sample, "[A-Za-z]+"))  %>%
  dplyr::mutate(Sample_Clean = stringr::str_replace(Sample, "RT.*", "R")) %>%
  dplyr::mutate(KRAS = str_replace(HGVSp_Short, "p.", "")) %>%
  dplyr::mutate(Source ="WES") %>%
  dplyr::select(Sample_Clean, KRAS, Source )  %>%
  dplyr::distinct() %>%
  dplyr::add_row(Sample_Clean ="123B", KRAS="WT", Source = "WES") %>% #If WT, not present in mutation data for WES
  dplyr::add_row(Sample_Clean ="80B", KRAS="WT", Source = "WES") #If WT, not present in mutation data for WES

kras_combined <- rbind(sanger_kras, wes_kras)
kras_combined <- kras_combined %>%
  dplyr::mutate(UniqueID = paste(Sample_Clean, KRAS, sep = "_")) %>% 
  tidyr::complete(Source,UniqueID) %>%
  dplyr::mutate(KRAS = ifelse(is.na(KRAS), "Absent", KRAS))

# Number of distinct mutations (excluding NA)
mutation_colors <- ArchR::paletteDiscrete(values = kras_combined$KRAS,set = "stallion", reverse = TRUE)
mutation_colors["Absent"] <- "gray" 

png(paste0(res_path,"/wes_sanger_comparison.png"), units = "in", res = 300, width = 8, height = 10)
ggplot(kras_combined, aes(x = Source, y = UniqueID, fill = KRAS)) +
  geom_tile() +
  scale_fill_manual(values = mutation_colors) +
  labs(x = "Source", y = "Sample_Mutation", fill = "Mutation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 7))
dev.off()
```

# Of the 34 samples with a TP53 mutation, export the mutation list 

```{r}
merged_maf@data  %>% dplyr::filter(Hugo_Symbol == "TP53") %>%
  dplyr::select(Tumor_Sample_Barcode, Hugo_Symbol, Reference_Allele, HGVSp, HGVSp_Short) %>%
  tidyr::separate(Tumor_Sample_Barcode, into = c("Patient", "Sample", "Treatment"), sep = "_", remove=FALSE) %>%
  write.csv(paste0(res_path, "TP53_mutations_list.csv"))

merged_maf@data  %>% dplyr::filter(Hugo_Symbol == "TP53") %>%
  dplyr::select(Tumor_Sample_Barcode, Hugo_Symbol, Reference_Allele, HGVSp, HGVSp_Short) %>%
  tidyr::separate(Tumor_Sample_Barcode, into = c("Patient", "Sample", "Treatment"), sep = "_", remove=FALSE) 
```
# Num gene category for all samples
```{r}
mut_data <- merged_maf@data %>%
  dplyr::mutate(patient_id = str_extract(Tumor_Sample_Barcode, "^PCA[0-9]+"))

mut_data_summary <- mut_data %>%
  dplyr::group_by(Hugo_Symbol) %>%
  dplyr::summarise(num_patients = dplyr::n_distinct(patient_id)) %>%
  dplyr::arrange(desc(num_patients)) 

mut_data_summary %>% write.csv(paste0(res_path, "num_patients_shared.csv"))

count_data <- mut_data_summary %>%
  dplyr::group_by(num_patients) %>%
  dplyr::summarise(num_genes = n()) %>%
  dplyr::arrange(num_patients)

png(paste0(res_path, "num_gene_category.png"), res=300, units = "in", width = 10, height=10)
ggplot(count_data, aes(x = factor(num_patients), y = num_genes, fill = num_genes)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label=num_genes), vjust=-0.2) +
  viridis::scale_fill_viridis() +
  labs(x = "Number of patients with mutated gene (out of 7 patients)", y = "Number of genes") +
  ggtitle("Number of genes for each category of patient count") + 
  theme_minimal()
dev.off()
```

# Num gene category for CM- samples
```{r}
mut_data_cm_neg <- merged_maf@data %>%
  dplyr::mutate(patient_id = str_extract(Tumor_Sample_Barcode, "^PCA[0-9]+")) %>%
  dplyr::filter(grepl(Tumor_Sample_Barcode, pattern="CM-"))

mut_data_summary_neg <- mut_data_cm_neg %>%
  dplyr::group_by(Hugo_Symbol) %>%
  dplyr::summarise(num_patients = dplyr::n_distinct(patient_id)) %>%
  dplyr::arrange(desc(num_patients)) 

mut_data_summary_neg %>% write.csv(paste0(res_path, "num_patients_shared_cm-.csv"))

count_data_cm_neg <- mut_data_summary_neg %>%
  dplyr::group_by(num_patients) %>%
  dplyr::summarise(num_genes = n()) %>%
  dplyr::arrange(num_patients)

png(paste0(res_path, "num_gene_category_cm-.png"), res=300, units = "in", width = 10, height=10)
ggplot(count_data_cm_neg, aes(x = factor(num_patients), y = num_genes, fill = num_genes)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label=num_genes), vjust=-0.2) +
  viridis::scale_fill_viridis() +
  labs(x = "Number of patients with mutated gene (out of 7 patients)", y = "Number of genes") +
  ggtitle("Number of genes for each category of patient count") + 
  theme_minimal()
dev.off()

mut_data_summary_neg
```


# Exploring sample summary statistics
```{r}
getGeneSummary(merged_maf)
merged_maf@gene.summary
getFields(merged_maf)
```

# Creating MAF summary plot
```{r}
png(paste0(res_path,"/mafsummary.png"), units = "in", res = 300, width = 8, height = 10)
plotmafSummary(maf = merged_maf, rmOutlier = TRUE, addStat = 'median', dashboard = TRUE)
dev.off()
```

# Oncoplot 

## Set up color palette

```{r}
n <- length(unique(merged_maf@clinical.data$Patient))
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
patient_colors <- sample(col_vector, n)
#patient_colors = RColorBrewer::brewer.pal(n =n,name = 'Spectral')
names(patient_colors) = unique(merged_maf@clinical.data$Patient)

treatment_colors = RColorBrewer::brewer.pal(n = length(unique(merged_maf@clinical.data$Treatment)),name = 'Accent')
names(treatment_colors) = unique(merged_maf@clinical.data$Treatment)
treatment_colors

sample_label_colors = c("#F22B6E", "#82BAB9")
names(sample_label_colors) = unique(merged_maf@clinical.data$Sample_Label)

stage_colors = RColorBrewer::brewer.pal(n = length(unique(merged_maf@clinical.data$Stage)),name = 'Dark2')
names(stage_colors) = unique(merged_maf@clinical.data$Stage)

anno_colors = list("Patient" = patient_colors,
  "Treatment" = treatment_colors,
  "Sample_Label" = sample_label_colors,
  "Stage" = stage_colors)
```

## Loading in cosmic genes

```{r}
data(cgc_67, package = "COSMIC.67")
cosmic_genes <- getGeneSummary(merged_maf)$Hugo_Symbol[getGeneSummary(merged_maf)$Hugo_Symbol %in% cgc_67$SYMBOL]
```

## Oncoplot for CM- for all genes (Patients only)

```{r}
cm_neg_samples <- merged_maf@clinical.data$Tumor_Sample_Barcode[grep(merged_maf@clinical.data$Tumor_Sample_Barcode, pattern="*CM-$")]
merged_maf_cm_neg <- subsetMaf(maf=merged_maf, tsb=cm_neg_samples)

tmb_df_cm_neg <- tmb(maf = merged_maf, captureSize=  36.5, logScale=FALSE) %>%
  dplyr::select(Tumor_Sample_Barcode, total_perMB) %>%
  dplyr::filter(Tumor_Sample_Barcode %in% cm_neg_samples) %>%
  as.data.frame()

png(paste0(res_path,"/oncoplot_snp_indel_cm-.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf_cm_neg, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, top = 50, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df_cm_neg, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.6, writeMatrix=TRUE)
dev.off()
```

## Oncoplot for CM- for COSMIC genes (Patients only)

```{r}
png(paste0(res_path,"/oncoplot_cosmic_snp_indel_cm-.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf_cm_neg, genes=cosmic_genes, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, top = 50, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df_cm_neg, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.6, writeMatrix=TRUE)
dev.off()
```

## Oncoplot for CM- with matching TISSUE for all genes (Patients only)

```{r}
cm_neg_tissue_samples <- merged_maf@clinical.data$Tumor_Sample_Barcode[grep(merged_maf@clinical.data$Tumor_Sample_Barcode, pattern="*CM-$|TISSUE$")]

# Extract base identifiers
base_ids <- str_extract(cm_neg_tissue_samples, ".*(?=_(CM-|TISSUE))")
base_ids
# Find unique identifiers with both CM- and TISSUE counterparts
unique_ids <- unique(base_ids)
paired_ids <- unique_ids[sapply(unique_ids, function(id) {
    any(grepl(paste0(id, "_CM-"), cm_neg_tissue_samples)) &&
    any(grepl(paste0(id, "_TISSUE"), cm_neg_tissue_samples))
})]

# Filter for CM- samples with a corresponding TISSUE sample
cm_neg_tissue_samples_final <- cm_neg_tissue_samples[base_ids %in% paired_ids & grepl("_CM-|TISSUE", cm_neg_tissue_samples)]
merged_maf_cm_neg_TISSUE <- subsetMaf(maf=merged_maf, tsb=cm_neg_tissue_samples_final)

tmb_df_cm_neg_TISSUE <- tmb(maf = merged_maf, captureSize=  36.5, logScale=FALSE) %>%
  dplyr::select(Tumor_Sample_Barcode, total_perMB) %>%
  dplyr::filter(Tumor_Sample_Barcode %in% cm_neg_tissue_samples_final) %>%
  as.data.frame()

png(paste0(res_path,"/oncoplot_snp_indel_cm-_TISSUE.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf_cm_neg_TISSUE, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, top = 50, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df_cm_neg_TISSUE, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.6, writeMatrix=TRUE, sampleOrder=cm_neg_tissue_samples_final)
dev.off()
```

## Oncoplot for CM- with matching TISSUE for COSMIC genes (Patients only)

```{r}
png(paste0(res_path,"/oncoplot_cosmic_snp_indel_cm-_TISSUE.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf_cm_neg_TISSUE, genes = cosmic_genes, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, top = 50, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df_cm_neg_TISSUE, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.6, writeMatrix=TRUE, sampleOrder=cm_neg_tissue_samples_final)
dev.off()
```
## Oncoplot for non-TISSUE PDO samples with all 4 conditions for all genes

```{r}
# Extract base identifiers
base_ids <- str_extract(merged_maf@clinical.data$Tumor_Sample_Barcode, ".*(?=_[^_]+$)")

# Unique base identifiers
unique_ids <- unique(base_ids)

# Check for each identifier if all four conditions are present
# Determine if all four conditions are present for each identifier
valid_ids <- unique_ids[sapply(unique_ids, function(id) {
    all(c(paste0(id, "_CM-"), paste0(id, "_CM+"), paste0(id, "_T1"), paste0(id, "_T2")) %in% merged_maf@clinical.data$Tumor_Sample_Barcode)
})]

# Select samples with valid identifiers
all_4_conditions_present_samples <- merged_maf@clinical.data$Tumor_Sample_Barcode[base_ids %in% valid_ids]
all_4_conditions_present_samples_no_TISSUE <- all_4_conditions_present_samples %>%
  grep(pattern=".*TISSUE", value = TRUE, invert=TRUE)
all_4_conditions_present_samples_no_TISSUE
merged_maf_all_4_conditions<- subsetMaf(maf=merged_maf, tsb=all_4_conditions_present_samples_no_TISSUE)

tmb_df_all_4_conditions <- tmb(maf = merged_maf_all_4_conditions, captureSize=  36.5, logScale=FALSE) %>%
  dplyr::select(Tumor_Sample_Barcode, total_perMB) %>%
  as.data.frame()

png(paste0(res_path,"/oncoplot_snp_indel_all4conditions.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf_all_4_conditions, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, top = 50, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df_all_4_conditions, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.6, writeMatrix=TRUE,
sampleOrder=all_4_conditions_present_samples_no_TISSUE)
dev.off()
```
## Oncoplot for non-TISSUE PDO samples with all 4 conditions for COSMIC genes

```{r}
png(paste0(res_path,"/oncoplot_cosmic_snp_indel_all4conditions.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf_all_4_conditions, genes=cosmic_genes, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, top = 50, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df_organoid_neg, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.6, writeMatrix=TRUE, sampleOrder=all_4_conditions_present_samples_no_TISSUE)
dev.off()
```

## Oncoplot for all samples with all genes

```{r}
tmb_df <- tmb(maf = merged_maf, captureSize=  36.5, logScale=FALSE) %>%
  dplyr::select(Tumor_Sample_Barcode, total_perMB) %>%
  as.data.frame()

# Saving oncoplot - not that useful as there are many genes involved
png(paste0(res_path,"/oncoplot_snp_indel.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, top = 50, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.6, writeMatrix=TRUE)
dev.off()
```

## Oncoplot of all samples with genes within COSMIC database

```{r}
png(paste0(res_path,"/oncoplot_cosmic_snp_indel.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf, , genes=cosmic_genes, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.5, writeMatrix=TRUE,
         rightBarLims)
dev.off()
```

# Calculate Tumor mutational burden (TMB) score alone
```{r}
# All PDO + Tissue
png(paste0(res_path,"/tmb_pdac.png"), units = "in", res = 300, width = 12, height = 12)
tmb(maf = merged_maf, captureSize=  36.5, logScale=FALSE)
dev.off()

# Tissue only
png(paste0(res_path,"/tmb_tissueOnly.png"), units = "in", res = 300, width = 12, height = 12)
tmb(maf=merged_maf_tissue_only, captureSize=36.5, logScale=FALSE)
dev.off()

# Organoid only
png(paste0(res_path,"/tmb_OrganoidOnly.png"), units = "in", res = 300, width = 12, height = 12)
tmb(maf=merged_maf_organoid_only, captureSize=36.5, logScale=FALSE)
dev.off()

```
# Titv plot 
```{r}
merged_maf.titv = titv(maf = merged_maf, plot = FALSE, useSyn = FALSE)

png(paste0(res_path,"/titv.png"), units = "in", res = 300, width = 12, height = 12)
plotTiTv(res = merged_maf.titv, showBarcodes=TRUE, textSize=0.6, plotNotch=TRUE)
dev.off()
```

# Lollipop Plot of specific genes
```{r}
genes_lollipop <- c("TP53", "KRAS", "SMAD4", "CDKN2A", "ARID1A", "ATM", "TTN")
lollipop_res <- res_path
for (gene in genes_lollipop){
  png(paste0(lollipop_res,gene,"_lollipop.png"), units = "in", res = 300, width = 8, height = 6)
  lollipopPlot(
  maf = merged_maf,
  labelPos="all",
  gene = gene,
  showMutationRate = TRUE)
  dev.off()
}
```

# plotVAF of key genes
```{r}
samples_data_tidy <- merged_maf@data %>%
  tidyr::separate(Tumor_Sample_Barcode, into = c("Patient", "Sample", "Treatment"), sep = "_", remove=FALSE) %>%
  dplyr::mutate(isTissue = ifelse(Treatment=="TISSUE","TISSUE","ORGANOID")) %>%
  dplyr::mutate(vaf = t_alt_count/t_depth)
samples_data_tidy_cosmic <- samples_data_tidy %>%
  dplyr::filter(Hugo_Symbol %in% cosmic_genes[1:12])

color_palette <- c("#08519C", "#3182BD", "#6BAED6", "#BDD7E7", "#FF0000")

png(paste0(res_path,"/vaf.png"), units = "in", res = 300, width = 14, height = 10)
ggplot(samples_data_tidy_cosmic, aes(x=Patient, y=vaf, fill=Treatment))+
  geom_col(position="dodge")+
  scale_fill_manual(values = color_palette) +
  facet_wrap(~Hugo_Symbol) +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=90))
dev.off()

```
# Rainfall plot
```{r}
rainfallPlot(maf = merged_maf,detectChangePoints = TRUE, pointSize = 0.4)
```

# Somatic interactions
```{r}
somaticInteractions(maf = merged_maf, top = 25, pvalue = c(0.05, 0.1))
```

# Oncodrive
```{r}
laml.sig = oncodrive(maf = merged_maf_cn, minMut = 5, pvalMethod = 'zscore')
plotOncodrive(res = laml.sig, fdrCutOff = 0.1, useFraction = TRUE, labelSize = 0.3, bubbleSize=0.5)
```

# Oncogenic dysregualted pathways
```{r}
OncogenicPathways(maf = merged_maf)
```
# Plot specific oncogenic pathways
```{r}
PlotOncogenicPathways(maf = merged_maf, pathways = "TP53")
```

# Signatures
```{r}
laml.tnm = trinucleotideMatrix(maf = merged_maf,  ref_genome = "BSgenome.Hsapiens.UCSC.hg38")
laml.sign = estimateSignatures(mat = laml.tnm, nTry = 20, pConstant=0.001)
plotCophenetic(res = laml.sign)
laml.sig = extractSignatures(mat = laml.tnm, n = 6, pConstant =0.001)
laml.og30.cosm = compareSignatures(nmfRes = laml.sig, sig_db = "legacy")
pheatmap::pheatmap(mat = laml.og30.cosm$cosine_similarities, cluster_rows = FALSE, main = "cosine similarity against validated signatures")
maftools::plotSignatures(nmfRes = laml.sig, title_size = 1.2, sig_db = "SBS")
getwd()
```


# Mutations in TP53 in CM- samples

```{r}
merged_maf@data %>% dplyr::filter(Hugo_Symbol == "TP53") %>% dplyr::select(Hugo_Symbol,Tumor_Sample_Barcode, HGVSp_Short) %>% dplyr::filter(grepl("CM-", Tumor_Sample_Barcode)) %>%
  dplyr::group_by(HGVSp_Short) %>% dplyr::arrange(HGVSp_Short) %>% write.csv(paste0(res_path, "TP53_CM-_mutations.csv"))
```

