---
title: "02B_oncoplot_CN_ascat"
output: html_document
date: "2023-09-11"
---

The purpose of this notebook is to create the copy number table as input for maftools. 
This requires extracting the 
CN information for the mutated genes from ASCAT.

```{r}
library(stringr)
library(purrr)
library(tidyr)
library(dplyr)
library(readr)
library(stringr)
library(ggplot2)
library(maftools)
library(ConsensusClusterPlus)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(biomaRt)
library(ArchR)

res_path <- "../results/combined_batches_wes/"
```

# Prepare CN table

```{r}
add_sample_info <- function(file_path){
  file <- readr::read_tsv(file_path, show_col_types = FALSE)
  file <- file %>%
    dplyr::mutate(sample_name = stringr::str_extract(basename(file_path), "^[^.]+")) %>%
    dplyr::mutate(patient_id = str_extract(sample_name, "^[0-9]+"),
         treatment = str_extract(sample_name, "(?<=_)[^_]+(?=_)"),
         sample_label = case_when(
          str_detect(sample_name, "^([0-9]+B)") ~ "B",
          str_detect(sample_name, "^([0-9]+RT)") ~ "RT",
          TRUE ~ "None"))
  return(file)
}

# Modified function to read each file and convert WGD to numeric
read_file <- function(file_path) {
  # Try to read the file and handle any errors
  tryCatch({
    df <- read_delim(file_path, delim = "\t", escape_double = FALSE, col_types = cols(), trim_ws = TRUE)
    
    # Check if 'WGD' column exists and is of type character, then convert it to numeric
    if ("WGD" %in% names(df) && is.character(df$WGD)) {
      df$WGD <- as.numeric(df$WGD)
    }

    # Ensure the output is a data frame
    if (!is.data.frame(df)) {
      stop("The output is not a data frame.")
    }

    return(df)
  }, error = function(e) {
    message("Error reading file: ", file_path, "\nError: ", e$message)
    return(NULL) # Return NULL if there's an error
  })
}
```

```{r}
# Listing file paths
cnvs_paths <- list.files(paste0(res_path, "ascat/"),  full.names=TRUE, recursive=TRUE) %>%
  grep(pattern="\\cnvs.txt$", value =TRUE)

metrics_paths <- list.files(paste0(res_path, "ascat/"), full.names=TRUE, recursive=TRUE) %>%
  grep(pattern="\\metrics.txt$", value =TRUE)

purityploidy_paths <- list.files(paste0(res_path, "ascat/"), full.names=TRUE, recursive=TRUE) %>%
  grep(pattern="\\purityploidy.txt$", value =TRUE)

segments_paths <- list.files(paste0(res_path, "ascat/"), full.names=TRUE, recursive=TRUE) %>%
  grep(pattern="\\ments.txt$", value =TRUE)
segments_paths
# Reading in dataframes
cnvs <- purrr::map_dfr(cnvs_paths, add_sample_info)
metrics <- purrr::map_dfr(metrics_paths,  read_file, .id = "file_id")
purityploidy <- purrr::map_dfr(purityploidy_paths, add_sample_info)
segments <- purrr::map_dfr(segments_paths, add_sample_info)
```

# Defining the CN categories
```{r}
# Create variables for TCN and segment size
df <- segments %>%
  dplyr::mutate(TCN = nMajor + nMinor,
         segment_size_kb = (endpos - startpos) / 1000,
         heterozygosity = case_when(
           nMajor > 0 & nMinor > 0 ~ "heterozygous",
           nMajor > 0 & nMinor == 0 ~ "LOH",
           nMajor == 0 & nMinor > 0  ~ "LOH",
           nMajor == 0 & nMinor == 0 ~ "homozygous deletion"
         ),
         TCN_category = case_when(
           TCN == 0 ~ "homozygous deletion",
           TCN == 1 ~ "deletion leading to LOH",
           TCN == 2 ~ "wild type",
           TCN %in% 3:4 ~ "minor gain",
           TCN %in% 5:8 ~ "moderate gain",
           TCN >= 9 ~ "high-level amplification"
         ),
         size_category = case_when(
           segment_size_kb <= 100 ~ "0-100kb",
           segment_size_kb > 100 & segment_size_kb <= 1000 ~ "100kb-1Mb",
           segment_size_kb > 1000 & segment_size_kb <= 10000 ~ "1Mb-10Mb",
           segment_size_kb > 10000 & segment_size_kb <= 40000 ~ "10Mb-40Mb",
           segment_size_kb > 40000 ~ ">40Mb"
         ))

# Count occurrences of each combination of categories
df_counts <- df %>%
  dplyr::group_by(sample,heterozygosity, TCN_category, size_category) %>%
  dplyr::summarize(n = n(), .groups = "drop")

# Create complete set of all combinations of categories
all_combinations <- expand_grid(sample = unique(df_counts$sample),
  heterozygosity = c("heterozygous", "LOH", "homozygous deletion"),
                                TCN_category = c("homozygous deletion",
                                                 "deletion leading to LOH",
                                                 "wild type",
                                                 "minor gain",
                                                 "moderate gain",
                                                 "high-level amplification"),
                                size_category = unique(df$size_category)) %>%
  dplyr::filter(!(heterozygosity == "homozygous deletion" & TCN_category != "homozygous deletion" |
                  heterozygosity == "LOH" & TCN_category == "homozygous deletion" |
                  heterozygosity == "heterozygous" & TCN_category %in% c("homozygous deletion", 
                                                                         "deletion leading to LOH")))


# Join actual counts with all combinations
df_final <- left_join(all_combinations, df_counts, by = c("heterozygosity", "TCN_category", "size_category", "sample"))

df_final$n[is.na(df_final$n)] <- 0

df_final <- df_final %>%
  dplyr::mutate(TCN = case_when(
           TCN_category == "homozygous deletion" ~ "0",
           TCN_category == "deletion leading to LOH" ~ "1",
           TCN_category == "wild type" ~ "2",
           TCN_category == "minor gain" ~ "3-4",
           TCN_category == "moderate gain"~ "5-8",
           TCN_category == "high-level amplification" ~ "9"
         ))%>%
    dplyr::mutate(patient_id = str_extract(sample, "^[0-9]+"),
         treatment = str_extract(sample, "(?<=_)[^_]+(?=_)"),
         sample_label = case_when(
          str_detect(sample, "^([0-9]+B)") ~ "B",
          str_detect(sample, "^([0-9]+RT)") ~ "RT",
          TRUE ~ "None"),
         sample_name_clean = paste(patient_id, sample_label, treatment, sep="_"))
```
# Create granges object from CN categories file
```{r}
cn_granges <- makeGRangesFromDataFrame(df, start.field="startpos", end.field="endpos", keep.extra.columns=TRUE)
```
# Get coordinate information of mutated genes
```{r}
merged_maf <- read.maf("../results/combined_batches_wes/merged_mafs_allBatches_160124_maftools.maf", 
                 verbose=TRUE)

my_genes = getGeneSummary(merged_maf)$Hugo_Symbol
m <- useMart('ensembl', dataset='hsapiens_gene_ensembl') # create a mart object
my_genes_info<- getBM(mart=m, attributes=c('hgnc_symbol', 'description', 'chromosome_name',
                                 'start_position', 'end_position', 'strand',
                                 'ensembl_gene_id'),
            filters='hgnc_symbol', values=my_genes) 

my_genes_granges <-my_genes_info %>%
  dplyr::mutate(strand_rename = ifelse(strand==-1, "-", ifelse(strand==1, "+", "*"))) %>%
  dplyr::select(-strand) %>%
  makeGRangesFromDataFrame(start.field="start_position",
                                             end.field="end_position",
                           strand.field="strand_rename",
                                             keep.extra.columns = TRUE)
```
# Find overlaps
```{r}
hits <- findOverlaps(cn_granges, my_genes_granges)
```

# Create new dataframe with the gene information

```{r}
process_hit <- function(hit){
  # Extract the gene name 
  overlapping_gene <- my_genes_info$hgnc_symbol[subjectHits(hit)]

  # Extract the CN region
  df_hit <- df[queryHits(hit), ]
  
  # Add a new column for overlapping gene
  df_hit$Gene <- overlapping_gene
  
  return(df_hit)
}
# takes a while
list_of_hits <- split(hits, seq_along(hits))
# takes a while too
result_df <- purrr::map_df(list_of_hits, process_hit)

result_df_final <- result_df %>%
  dplyr::mutate(CN = case_when(
      grepl("deletion", TCN_category) ~ "Del",
      grepl("minor gain", TCN_category) ~ "Gain",
      grepl("moderate gain", TCN_category) ~ "Amp",
      grepl("wild type", TCN_category) ~ "wild type",
      TRUE ~ "NA" 
  ))
```

# Create the CN table

```{r}
# Need to wrangle smaple_name to be same format as the merged_maf file
CN_table <- result_df_final %>%
  dplyr::mutate(sample_name_reformat = paste(paste0("PCA", patient_id),
                                    paste0(patient_id, sample_label),
                                    treatment,
                                    sep="_")) %>%
  dplyr::filter(CN != "wild type") %>%
  dplyr::select(Gene, sample_name_reformat, CN)
length(unique(CN_table$sample_name_reformat))
CN_table
# Strange issue with sample name from 63RT to 63RT3, hence update the CN table
CN_table <- CN_table %>%
  dplyr::mutate(sample_name_reformat = stringr::str_replace(sample_name_reformat, "63RT", "63RT3"))
```

```{r}
data(cgc_67, package = "COSMIC.67")
# Get cosmic genes
cosmic_genes <- getGeneSummary(merged_maf)$Hugo_Symbol[getGeneSummary(merged_maf)$Hugo_Symbol %in% cgc_67$SYMBOL]

# Data wrangling to remove samples from tissue + retain genes in COSMIC db only
data_matrix_cosmic_noTissue <-  CN_table %>%
  dplyr::filter(Gene %in% cosmic_genes) %>%
  pivot_wider(names_from= sample_name_reformat, values_from= CN, values_fn = list, values_fill = list("Neutral")) %>%
  tibble::column_to_rownames(var = "Gene") %>%
  dplyr::select(-matches("TISSUE$")) %>%
  as.matrix()

CN_table

data_matrix_cosmic_noTissue_simple <- matrix(sapply(data_matrix_cosmic_noTissue, `[[`, 1), nrow=nrow(data_matrix_cosmic_noTissue), ncol=ncol(data_matrix_cosmic_noTissue))
rownames(data_matrix_cosmic_noTissue_simple) <- rownames(data_matrix_cosmic_noTissue)
colnames(data_matrix_cosmic_noTissue_simple) <- colnames(data_matrix_cosmic_noTissue)

#Create a color mapping for the CN alteration statuses
cn_col_mapping <- c(
  Amp = "#ca472f",
  Gain = "#f6c85f",
  Del = "#0b84a5",
  Neutral = "#F5F5F5"
)

# Calculate the number of non-"neutral" entries for each row i.e. # samples per gene with a CNA
neutral_count <- rowSums(data_matrix_cosmic_noTissue_simple != "Neutral")

# Order the rows based on the number of non-neutral entries, with gene with highest CNA at top
#ordered_row_indices <- order(neutral_count, decreasing = TRUE)
#data_matrix_cosmic_noTissue_simple <- data_matrix_cosmic_noTissue_simple[ordered_row_indices, ]
```

# Group samples by condition
```{r}
cm_neg_samples <- grep(colnames(data_matrix_cosmic_noTissue_simple),pattern="CM-", value=TRUE)
t1_samples <- grep(colnames(data_matrix_cosmic_noTissue_simple),pattern="T1", value=TRUE)
cm_pos_samples <- grep(colnames(data_matrix_cosmic_noTissue_simple),pattern="CM\\+", value=TRUE)
t2_samples <- grep(colnames(data_matrix_cosmic_noTissue_simple),pattern="T2", value=TRUE)

data_matrix_cosmic_noTissue_simple_fewGenesOnly <- data_matrix_cosmic_noTissue_simple[c("KRAS", "TP53", "SMAD4", "CDKN2A", "ATM"),c(cm_neg_samples, t1_samples, cm_pos_samples, t2_samples)]

data_matrix_cosmic_noTissue_simple_KRAS_only <- data_matrix_cosmic_noTissue_simple[c("KRAS"),c(cm_neg_samples, t1_samples, cm_pos_samples, t2_samples)]

data_matrix_cosmic_noTissue_simple_KRAS_only_normalOrder <- data_matrix_cosmic_noTissue_simple[c("KRAS"),]

#  Create the heatmap using ComplexHeatmap
png(paste0(res_path,"/complexheatmap_CN_group_condition.png"), units = "in", res = 300, width = 12, height = 17)
ComplexHeatmap::Heatmap(data_matrix_cosmic_noTissue_simple_fewGenesOnly, 
        name = "CN alterations", 
        col = cn_col_mapping,
        show_row_names = TRUE, 
        show_column_names = TRUE, 
        cluster_rows = FALSE, 
        cluster_columns = TRUE,
        row_names_gp = gpar(fontsize = 12),
        column_names_gp = gpar(fontsize = 12),
        rect_gp = gpar(col = "black", lwd = 0.2))
dev.off()

#  Create the heatmap using ComplexHeatmap
png(paste0(res_path,"/complexheatmap_CN_group_condition_KRAS_only.png"), units = "in", res = 300, width = 12, height = 17)
ComplexHeatmap::Heatmap(data_matrix_cosmic_noTissue_simple_KRAS_only, 
        name = "CN alterations", 
        col = cn_col_mapping,
        show_row_names = TRUE, 
        show_column_names = TRUE, 
        cluster_rows = FALSE, 
        cluster_columns = TRUE,
        row_names_gp = gpar(fontsize = 12),
        column_names_gp = gpar(fontsize = 12),
        rect_gp = gpar(col = "black", lwd = 0.2))
dev.off()

png(paste0(res_path,"/complexheatmap_CN_KRAS_only.png"), units = "in", res = 300, width = 12, height = 17)
ComplexHeatmap::Heatmap(data_matrix_cosmic_noTissue_simple_KRAS_only_normalOrder, 
        name = "CN alterations", 
        col = cn_col_mapping,
        show_row_names = TRUE, 
        show_column_names = TRUE, 
        cluster_rows = FALSE, 
        cluster_columns = TRUE,
        row_names_gp = gpar(fontsize = 12),
        column_names_gp = gpar(fontsize = 12),
        rect_gp = gpar(col = "black", lwd = 0.2))
dev.off()
```

