---
title: "Untitled"
output: html_document
date: "2024-01-16"
---

```{r}
library(stringr)
library(tidyr)
library(dplyr)
library(readr)
library(stringr)
library(ggplot2)
library(maftools)
library(ConsensusClusterPlus)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(biomaRt)
library(ArchR)

res_path <- "../results/combined_batches_wes/"
```

# Load in maf file, pass in cnTable, and process

```{r}
merged_maf_cn <- read.maf("../results/combined_batches_wes/merged_mafs_allBatches_160124_maftools.maf",
                  cnTable = CN_table,
                  verbose=TRUE)

# Load in metadata containing KRAS mutation information from sanger seuqencing
metadata_wes <- readxl::read_excel("../metadata/metadata_all_samples.xlsx")

# Cleaning the clinical data column to match with metadata spec
merged_maf_cn@clinical.data <- merged_maf_cn@clinical.data %>%
  tidyr::separate(Tumor_Sample_Barcode, into = c("Patient", "Sample", "Treatment"), sep = "_", remove=FALSE) %>%
  dplyr::mutate(Sample_Label = stringr::str_extract(Sample, "[A-Za-z]+")) %>%
  dplyr::mutate(Sample_Clean = stringr::str_replace(Sample, "RT.*", "R")) %>%
  data.table::setDT() 

# Adding in metadata info
merged_maf_cn@clinical.data <- dplyr::left_join(merged_maf_cn@clinical.data, metadata_wes, by=c("Sample_Clean" = "Label")) %>%
  dplyr::rename(Stage = `Stage at diagnosis...10`)

tissue_samples <- merged_maf_cn@clinical.data$Tumor_Sample_Barcode[grep(merged_maf_cn@clinical.data$Tumor_Sample_Barcode, pattern="*TISSUE$")]
organoid_samples <- merged_maf_cn@clinical.data$Tumor_Sample_Barcode[-grep(merged_maf_cn@clinical.data$Tumor_Sample_Barcode, pattern="*TISSUE$")]

merged_maf_tissue_only <- subsetMaf(merged_maf_cn, tsb= tissue_samples)
merged_maf_organoid_only <- subsetMaf(merged_maf_cn, tsb= organoid_samples)
```

# Setting up oncoplot colors

```{r}
# Setting up colors for variant classification 
vc_cols = ArchR::ArchRPalettes$circus
names(vc_cols) = c(
  'Frame_Shift_Del',
  'Missense_Mutation',
  'Nonsense_Mutation',
  'Multi_Hit',
  'Frame_Shift_Ins',
  'In_Frame_Ins',
  'Splice_Site',
  'In_Frame_Del',
  'Del',
  'Gain',
  'Amp',
  'Complex_Event'
)
vc_cn_cols <- c(
  'Del',
  'Gain',
  'Amp'
)

# Setting up annotation colors for clinical features
n_patients <- length(unique(merged_maf_cn@clinical.data$Patient))
merged_maf_cn@clinical.data
# Create a color ramp function based on the 'Spectral' palette
spectral_ramp <- colorRampPalette(brewer.pal(11, "Spectral"))
patient_colors <- spectral_ramp(n_patients)
names(patient_colors) = unique(merged_maf_cn@clinical.data$Patient)

treatment_colors = RColorBrewer::brewer.pal(n = length(unique(merged_maf_cn@clinical.data$Treatment)),name = 'Accent')
names(treatment_colors) = unique(merged_maf_cn@clinical.data$Treatment)

sample_label_colors = c("#F22B6E", "#82BAB9")
names(sample_label_colors) = unique(merged_maf_cn@clinical.data$Sample_Label)

stage_colors = RColorBrewer::brewer.pal(n = length(unique(merged_maf_cn@clinical.data$Stage)),name = 'Dark2')
names(stage_colors) = unique(merged_maf_cn@clinical.data$Stage)

anno_colors = list("Patient" = patient_colors,
  "Treatment" = treatment_colors,
  "Sample_Label" = sample_label_colors,
  "Stage" = stage_colors)
```
## Loading in cosmic genes

```{r}
data(cgc_67, package = "COSMIC.67")
cosmic_genes <- getGeneSummary(merged_maf)$Hugo_Symbol[getGeneSummary(merged_maf)$Hugo_Symbol %in% cgc_67$SYMBOL]
```

## Oncoplot for CM- for all genes (Patients only)

```{r}
cm_neg_samples <- merged_maf_cn@clinical.data$Tumor_Sample_Barcode[grep(merged_maf_cn@clinical.data$Tumor_Sample_Barcode, pattern="*CM-$")]
merged_maf_cm_neg <- subsetMaf(maf=merged_maf_cn, tsb=cm_neg_samples)

tmb_df_cm_neg <- tmb(maf = merged_maf, captureSize=  36.5, logScale=FALSE) %>%
  dplyr::select(Tumor_Sample_Barcode, total_perMB) %>%
  dplyr::filter(Tumor_Sample_Barcode %in% cm_neg_samples) %>%
  as.data.frame()

png(paste0(res_path,"oncoplots_cn/oncoplot_cn_cm-.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf_cm_neg, colors=vc_cols, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, top = 50, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df_cm_neg, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.6, writeMatrix=TRUE)
dev.off()
```

## Oncoplot for CM- for COSMIC genes (Patients only)

```{r}
png(paste0(res_path,"oncoplots_cn/oncoplot_cosmic_cn_cm-.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf_cm_neg, genes=cosmic_genes, colors=vc_cols, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, top = 50, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df_cm_neg, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.6, writeMatrix=TRUE)
dev.off()
```

## Oncoplot for CM- with matching TISSUE for all genes (Patients only)

```{r}
cm_neg_tissue_samples <- merged_maf_cn@clinical.data$Tumor_Sample_Barcode[grep(merged_maf_cn@clinical.data$Tumor_Sample_Barcode, pattern="*CM-$|TISSUE$")]

# Extract base identifiers
base_ids <- str_extract(cm_neg_tissue_samples, ".*(?=_(CM-|TISSUE))")

# Find unique identifiers with both CM- and TISSUE counterparts
unique_ids <- unique(base_ids)
paired_ids <- unique_ids[sapply(unique_ids, function(id) {
    any(grepl(paste0(id, "_CM-"), cm_neg_tissue_samples)) &&
    any(grepl(paste0(id, "_TISSUE"), cm_neg_tissue_samples))
})]

# Filter for CM- samples with a corresponding TISSUE sample
cm_neg_tissue_samples_final <- cm_neg_tissue_samples[base_ids %in% paired_ids & grepl("_CM-|TISSUE", cm_neg_tissue_samples)]
merged_maf_cm_neg_TISSUE <- subsetMaf(maf=merged_maf_cn, tsb=cm_neg_tissue_samples_final)

tmb_df_cm_neg_TISSUE <- tmb(maf = merged_maf, captureSize=  36.5, logScale=FALSE) %>%
  dplyr::select(Tumor_Sample_Barcode, total_perMB) %>%
  dplyr::filter(Tumor_Sample_Barcode %in% cm_neg_tissue_samples_final) %>%
  as.data.frame()

png(paste0(res_path,"oncoplots_cn/oncoplot_cn_cm-_TISSUE.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf_cm_neg_TISSUE, colors=vc_cols, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, top = 50, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df_cm_neg_TISSUE, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.6, writeMatrix=TRUE, sampleOrder=cm_neg_tissue_samples_final)
dev.off()
```

## Oncoplot for CM- with matching TISSUE for COSMIC genes (Patients only)

```{r}
png(paste0(res_path,"oncoplots_cn/oncoplot_cosmic_cn_cm-_TISSUE.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf_cm_neg_TISSUE, colors=vc_cols, genes = cosmic_genes, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, top = 50, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df_cm_neg_TISSUE, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.6, writeMatrix=TRUE, sampleOrder=cm_neg_tissue_samples_final)
dev.off()
```
## Oncoplot for non-TISSUE PDO samples with all 4 conditions for all genes

```{r}
# Extract base identifiers
base_ids <- str_extract(merged_maf_cn@clinical.data$Tumor_Sample_Barcode, ".*(?=_[^_]+$)")

# Unique base identifiers
unique_ids <- unique(base_ids)

# Check for each identifier if all four conditions are present
# Determine if all four conditions are present for each identifier
valid_ids <- unique_ids[sapply(unique_ids, function(id) {
    all(c(paste0(id, "_CM-"), paste0(id, "_CM+"), paste0(id, "_T1"), paste0(id, "_T2")) %in% merged_maf_cn@clinical.data$Tumor_Sample_Barcode)
})]

# Select samples with valid identifiers
all_4_conditions_present_samples <- merged_maf_cn@clinical.data$Tumor_Sample_Barcode[base_ids %in% valid_ids]
all_4_conditions_present_samples_no_TISSUE <- all_4_conditions_present_samples %>%
  grep(pattern=".*TISSUE", value = TRUE, invert=TRUE)

merged_maf_all_4_conditions<- subsetMaf(maf=merged_maf_cn, tsb=all_4_conditions_present_samples_no_TISSUE)

tmb_df_all_4_conditions <- tmb(maf = merged_maf, captureSize=  36.5, logScale=FALSE) %>%
  dplyr::select(Tumor_Sample_Barcode, total_perMB) %>%
  as.data.frame()

png(paste0(res_path,"oncoplots_cn/oncoplot_cn_all4conditions.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf_all_4_conditions, colors=vc_cols, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, top = 50, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df_all_4_conditions, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.6, writeMatrix=TRUE,
sampleOrder=all_4_conditions_present_samples_no_TISSUE)
dev.off()
```
## Oncoplot for non-TISSUE PDO samples with all 4 conditions for COSMIC genes

```{r}
png(paste0(res_path,"oncoplots_cn/oncoplot_cosmic_cn_all4conditions.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf_all_4_conditions, genes=cosmic_genes, colors=vc_cols, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, top = 50, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df_organoid_neg, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.6, writeMatrix=TRUE, sampleOrder=all_4_conditions_present_samples_no_TISSUE)
dev.off()
```

## Oncoplot for all samples with all genes

```{r}
tmb_df <- tmb(maf = merged_maf, captureSize=  36.5, logScale=FALSE) %>%
  dplyr::select(Tumor_Sample_Barcode, total_perMB) %>%
  as.data.frame()

# Saving oncoplot - not that useful as there are many genes involved
png(paste0(res_path,"oncoplots_cn/oncoplot_cn.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf_cn, colors=vc_cols, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, top = 50, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.6, writeMatrix=TRUE)
dev.off()
```

## Oncoplot of all samples with genes within COSMIC database

```{r}
png(paste0(res_path,"oncoplots_cn/oncoplot_cosmic_cn.png"), units = "in", res = 300, width = 12, height = 17)
oncoplot(maf = merged_maf_cn, colors=vc_cols, genes=cosmic_genes, clinicalFeatures=c("Patient","Treatment","Sample_Label","Stage"), annotationColor=anno_colors, sortByAnnotation=TRUE, anno_height= 1, showTumorSampleBarcodes=TRUE, barcodeSrt = 90,  barcode_mar=10, showTitle=FALSE, drawBox=TRUE, gene_mar=4.5, draw_titv=FALSE, annotationFontSize=1, topBarData = tmb_df, drawColBar=TRUE, legendFontSize =1, SampleNamefontSize=1, fontSize = 0.5, writeMatrix=TRUE,
         rightBarLims)
dev.off()
```
